---
title: "Parameter Estimation and Machine Learning"
author: "Yue Shi, Ph.D candidate, University of Washington"
date: "5/8/2018"
output:
  html_document:
    toc: yes
    toc_depth: '5'
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Maximum Likelihood Estimation (MLE)

Likelihood is: 
$$
L(\theta:Data)=P(Data | \theta)
$$
Maximum likelihood Estimation is to find $\theta^*$ to maximize the likelihood. This $\theta^*$ is the maximum likelihood estimate of $\theta$. 

#### Example: Halitosis example
Halitosis, colloquially called bad breath, is a symptom in which a noticeably unpleasant odor is present on the exhaled breath. Halitosis is partly genetically determined. The genotype aa has a 40% chance of getting the disease, and the other two possible genotypes, AA and Aa, each has a 10% chance of getting the disease. We want to estimate the frequency of the A allele.

If the gene frequency of the  A  allele is p, and that of a is 1-p, then the frequency of the disease is expected to be (if the genotypes are in Hardy-Weinberg proportions as a result of random mating):
  
  $$F = p^2 \times (0.1) + 2p(1-p) \times (0.1) + (1-p)^2 \times (0.4)$$
  
Now suppose we observe 1000 individuals and find that the 182 of them have the disease.
Using a binomial distribution, the probability that 182 out of 1000
have the disease is the binomial class probability for 182 out of 1000
when the probability of the event is F (which is a function of p).  This is

$$\frac{1000!}{182!~818!} F^{182} (1-F)^{818}$$

```{r}
##Given p, calcualte the probability of getting Halitosis
H=function(x){
  hal=0.1*x^2 + 0.1*2*x*(1-x) + 0.4*(1-x)^2 
  return(hal)
}

##Given H, calcualte the log likelihood of getting the data. 
LD=function(x){
  LL=182*log(x)+818*log(1-x)
  return(LL)
}

##Now let's find p which maximize the LD. 
fp=seq(0,1,0.001)
fLD=LD(H(fp))
mle=fp[which.max(fLD)]
mle
plot(fLD~fp, xlab="Allele frequency of A", ylab="Data Likelihood")
grid(10,10)
abline(v=mle,col="red")
```

## Maximum a Posteriori (MAP) Estimation

Posterior probability is: 
$$
P(\theta | D) = \frac{P(D|\theta)P(\theta)}{P(D)} = \frac{P(D|\theta)P(\theta)}{\int P(D|\theta)P(\theta)d\theta}
$$
Since the demoninator is not a function of $\theta$, therefore we can ignore the denominator， then
$$
\begin{eqnarray}
P(\theta | D) &\propto& P(D|\theta)P(\theta)\\
Posterior &\propto& Likelihood * Prior
\end{eqnarray}
$$
MAP estimation is to find $\theta$ that maximizes posterior $P(\theta|D)$.  
The difference between MLE and MAP is that:  
For MLE: find $\theta$ that maximizes $logP(D|\theta)$; whereas  
For MAP: find $\theta$ that maximizes $logP(D|\theta) + logP(\theta)$. 

The difference between MAP and Bayesian estimation is that:  
MAP ignore the demoninator which integrate all of the possible $\theta$, which makes Bayesian method much more computational demanding. MAP is must faster and popular now. 

#### beta distribution

**beta distribution**is parameterized by two shape constraints $\alpha$ and $\beta$. It does the job nicely for expressing the prior belief of probability which is restricted in the range [0,1]. 

$$
P(p)=\frac{1}{B(\alpha,\beta)}p^{\alpha-1}(1-p)^{\beta-1}
$$
where $B(\alpha,\beta) is the beta function.  

When both $\alpha$ and $\beta$ are greater than zero, it has the following properties:
$$
mode=\frac{\alpha-1}{\alpha+\beta-2}
$$
$$
variance=\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
$$

#### Example:Thumbtack example (unfair coin)
```{r}
nh <- 100 #the number of heads
nt <- 50 #the number of tails
alpha <- 1000 #alpha and beta are hyperparameters
beta <- 1000
logLikelihood <- function(p, nh, nt){
  return(nh*log(p)+nt*log(1-p))
}

logPosterior <- function(p, nh, nt, alpha, beta){
  return((nh+alpha-1)*log(p)+(nt+beta-1)*log(1-p))
}
p=seq(0,1,0.01)
LL=logLikelihood(p,nh,nt)
LP=logPosterior(p,nh,nt,alpha,beta)
par(mfrow=c(1,2))
plot(LL~p, ylab="Likelihood",main="Likelihood")
plot(LP~p, ylab="Likelihood",main="Posterior (alpha=1000, beta=1000)")
LL.max=p[which.max(LL)]
LL.max
nh/(nh+nt)
LP.max=p[which.max(LP)]
LP.max
(nh+alpha-1)/(nh+nt+alpha+beta-1)
```

Let's minimize the influence of priors
```{r}
alpha=30
beta=30
LL=logLikelihood(p,nh,nt)
LP=logPosterior(p,nh,nt,alpha,beta)
par(mfrow=c(1,2))
plot(LL~p, ylab="Likelihood",main="Likelihood")
plot(LP~p, ylab="Likelihood",main="Posterior (alpha=30, beta=30)")
LL.max=p[which.max(LL)]
LL.max
nh/(nh+nt)
LP.max=p[which.max(LP)]
LP.max
(nh+alpha-1)/(nh+nt+alpha+beta-1)
```

